---
title: "Project: Volatility, VaR, Expected Shortfall & Portfolio Optimization"
date: "Cuauhtemoc Daniel Puente Cavazos "
output: pdf_document
urlcolor: blue
geometry: left=0.75in, right=0.75in, top=0.75in, bottom=0.75in
---

```{r, echo=FALSE}
if (!require("xts")) install.packages("xts")
library(xts)

if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)


```


Load packages
```{r, message=FALSE, warning=FALSE}
options(repos = c(CRAN="http://cran.rstudio.com")) 
if (!require("MASS")) install.packages("MASS")
if (!require("forecast")) install.packages("forecast")
if (!require("rugarch")) install.packages("rugarch") 
if (!require("rmgarch")) install.packages("rmgarch") 
if (!require("timetk")) install.packages("timetk")
if (!require("quantmod")) install.packages("quantmod") 
library(forecast); library(MASS); library(rugarch); library(rmgarch); library(quantmod); library(graphics)
library(xts); library(timetk); library(stats); library(ggplot2); library(ggfortify); library(ggpubr)
```

# 1. Univariate Volatility

## 1.1 Data

S&P500 index ([Source](https://finance.yahoo.com/quote/^GSPC/)), for plots see file "9a. Volatility"
```{r, message=FALSE, warning=FALSE}
NASDAQ <- quantmod::getSymbols("^IXIC", src="yahoo", return.class="xts", from="2010-01-01", auto.assign=FALSE)

#print(colnames(NASDAQ))
NASDAQ_close <- NASDAQ[, 4]  
colnames(NASDAQ_close) <- "NASDAQ"
NASDAQ_r <- na.omit(diff(log(NASDAQ_close))) * 100
tail(NASDAQ_close)

```

## 1.2 Univariate Volatility Model 
ARMA(1,0)-GARCH(1,1) model specification
```{r}
model <- rugarch::ugarchspec(
         variance.model = list(model="sGARCH", garchOrder=c(1, 1)), 
         mean.model = list(armaOrder=c(1, 0)))
```

 model fit
```{r}
model_fit <- rugarch::ugarchfit(spec=model, data=NASDAQ_r, solver='hybrid')
```

Forecast
```{r, warning=FALSE}
n_future <- 8
model_forecast <- rugarch::ugarchforecast(fit=model_fit, n.ahead=n_future)
mean_f <- as.numeric(model_forecast@forecast$seriesFor)/100 # from % to fraction
std_f <- as.numeric(model_forecast@forecast$sigmaFor)/100 # from % to fraction
```

## 1.3 Value at Risk

At $\alpha = 1\%$ for a $50,000 position in NASDAQ
```{r, fig.width=7, fig.height=3}
VaR_f <- 50000 * abs(mean_f - 2.33 * std_f)

```

Plot
```{r, fig.width=7, fig.height=3, warning=FALSE}
id <- timetk::tk_index(NASDAQ_r)
id_f <- timetk::tk_make_future_timeseries(id, length_out=n_future, inspect_weekdays=TRUE)

VaR_f <- xts(VaR_f, order.by=id_f)
colnames(VaR_f) <- "VaR"

ggplot(data=VaR_f, aes(x=index(VaR_f), y=VaR)) + 
  geom_line(color="deeppink4") + 
  geom_point(color="deeppink4") +
  labs(x="Date", y="", title = "Forecast of Value at Risk for a $50,000 position in NASDAQ") +
  theme_minimal() + scale_x_date(date_breaks="1 day", date_labels = "%b %d") +
  theme(plot.title = element_text(size=10)) 
```

## 1.4 Expected Shortfall

At $\alpha=1\%$ for a $50,000 position in NASDAQ
```{r}
ES_f <- 50000 * abs(mean_f - 2.64 * std_f)

```

Plot
```{r, fig.width=7, fig.height=3}
ES_f <- xts(ES_f, order.by=id_f)
colnames(ES_f) <- "ES"

ggplot(data=ES_f, aes(x=index(ES_f), y=ES)) + 
  geom_line(color="springgreen4") + 
  geom_point(color="springgreen4") +
  labs(x="Date", y="", title = "Forecast of Expected Shortfall for a $50,000 position in NASDAQ") +
  theme_minimal() + scale_x_date(date_breaks="1 day", date_labels = "%b %d") +
  theme(plot.title = element_text(size=10))
```

# 2. Multivariate Volatility

## 2.1 Data
THis come from, [tickers](https://finance.yahoo.com/lookup/)
```{r, message=FALSE, warning=FALSE}
names <- c("Pfizer", "AstraZeneca", "Johnson&Johnson", "Merck", "Eli Lilly", "Bristol-Myers Squibb")
tickers <- c("PFE", "AZN", "JNJ", "MRK", "LLY", "BMY")


color_palette <- c("lightblue3", "darkorange", "yellowgreen", "cyan3")
```

```{r}
findat <- quantmod::getSymbols(tickers, src="yahoo", from="2022-06-01")
prices_lst <- lapply(tickers, function(ticker) {
  get(ticker)[, paste0(ticker, ".Close")]
})
prices <- do.call(merge.xts, prices_lst)
colnames(prices) <- tickers
returns <- na.omit(diff(log(prices)))*100 # in % terms
```

Risk-free asset return, U.S. 3-month Treasury Bills ([Source](https://fred.stlouisfed.org/series/TB3MS))
```{r}
rf_rate <- 4.5 / 252

```

## 2.2 Multivariate Volatility Model

DCC model
```{r}
# Univariate GARCH model specification
uni_garch <- rugarch::ugarchspec(
             variance.model = list(model = 'sGARCH', garchOrder = c(1,1)),  
             mean.model = list(armaOrder = c(1,1)))

# Replicate the univariate model for each dimension
dim <- ncol(returns)
rep_garch <- rugarch::multispec(replicate(dim, uni_garch))

# Formulate the DCC model
DCC_model <- rmgarch::dccspec(uspec = rep_garch, dccOrder = c(1, 1), model = "DCC")

# Fit the model to the data
DCC_fit <- rmgarch::dccfit(spec = DCC_model, data = returns, solver = c("hybrid","solnp"))
```

forecast of the Covariance matrix, 1 step ahead
```{r}
fct <- rmgarch::dccforecast(DCC_fit, n.ahead=1)
S <- fct@mforecast$H[[1]]
S <- as.matrix(S[,,1])
S
```

Invert the forecasted Covariance matrix
```{r}
S_inv <- MASS::ginv(S)
S_inv
```

forecast of the mean, 1 step ahead
```{r}
mu <- fct@mforecast$mu
mu <- as.vector(mu)
mu
```

## 2.3 Optimal Tangency Portfolio Weights
```{r}
mu <- as.numeric(mu)

x_star <- (S_inv %*% (mu - rf_rate)) / sum(S_inv %*% (mu - rf_rate))
x_star
```

```{r}
x_star_pct <- x_star * 100
for (i in seq_along(names)) {
  cat(sprintf("%-24s: %.2f%%\n", names[i], x_star_pct[i]))
}

```
